---
title: "Partial Pooling"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Partial Pooling}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE
)
```

## Why partial pooling matters

In many conflict mortality applications you have a small number of regions (for example five), and the data quality may vary sharply across them.

A fully pooled model (one coefficient shared across regions) can be too rigid: it forces conflict to have the same effect everywhere.

A fully separate model (fit each region independently) can be too noisy: sparse regions can produce extreme estimates driven by reporting artefacts.

**Partial pooling** is the compromise: each region gets its own effect, but regions share information through a hierarchical prior.

`vrcmort` uses partial pooling in several places by default (for example region random intercepts), and also allows partial pooling for conflict effects.

## Region random intercepts

Both the mortality and reporting submodels include region-level random intercepts by cause group.

For mortality:

\[
 u^{(\lambda)}_{r,g} \sim \mathcal{N}(0, \sigma^{(\lambda)}_{u,g}).
\]

For reporting:

\[
 u^{(\rho)}_{r,g} \sim \mathcal{N}(0, \sigma^{(\rho)}_{u,g}).
\]

These terms capture persistent differences between regions (baseline death rates and baseline completeness) that are not explained by covariates.

## Partial pooling for conflict effects (random slopes)

A core extension in `vrcmort` is allowing the conflict effect to vary by region.

### Mortality side

Instead of a single conflict effect $\beta_{conf,g}$, we allow:

\[
\beta_{conf,r,g} = \beta_{conf,g} + b^{(\beta)}_{r,g},
\]

where

\[
 b^{(\beta)}_{r,g} \sim \mathcal{N}(0, \sigma_{\beta,g}).
\]

This lets conflict have a stronger association with true mortality in some regions than others, while shrinking weakly informed regions back towards the global mean.

In the Stan program the region-specific effects are parameterised as a non-centred random effect for better sampling.

### Reporting side

Similarly, the effect of conflict on reporting completeness can vary by region:

\[
\gamma_{conf,r,g} = \gamma_{conf,g} + b^{(\gamma)}_{r,g}, \quad
b^{(\gamma)}_{r,g} \sim \mathcal{N}(0, \sigma_{\gamma,g}).
\]

This is useful when some regions experience a much sharper collapse in registration after conflict starts.

## Region-specific time trends

The base model includes national random walks in time for both mortality and reporting.

In many conflict settings, you also want region-specific departures from the national trend. `vrcmort` supports this by adding a region-specific random walk deviation around the national time effect.

Conceptually:

\[
 v^{(\lambda)}_{t,g} + v^{(\lambda,reg)}_{r,t,g}
\]

and

\[
 v^{(\rho)}_{t,g} + v^{(\rho,reg)}_{r,t,g}.
\]

These terms can absorb local shocks (for example temporary outages or spikes) without forcing all regions to move together.

## How to specify pooling in vrcmort

Pooling options are controlled through the component specifications passed to `vrcm()`.

```{r, eval=FALSE}
library(vrcmort)

# Region-varying conflict effects and region-specific time trends
mort_pp <- vrc_mortality(~ facility, conflict = "region", time = "region")
rep_pp  <- vrc_reporting(~ facility, conflict = "region", time = "region")

fit_pp <- vrcm(
  mortality = mort_pp,
  reporting = rep_pp,
  data = vr_long,
  t0 = conflict_start_month,
  chains = 4,
  iter = 1000
)

# Extract region-specific conflict effects
vrc_conflict_effects(fit_pp, component = "mortality")
vrc_conflict_effects(fit_pp, component = "reporting")
```

## Practical advice

- If you have only a few regions, partial pooling is usually safer than fitting separate models.
- If reporting collapse clearly differs by region, allowing region-varying conflict effects in the reporting model is often important.
- Region-specific time trends can be valuable, but they add parameters. If sampling becomes difficult, start with national trends only and add region-specific trends later.
